# A comparison of indexing and subsetting in R and Python {-}


## Dataframes {-}

In what follows we assume Python has access to a dataframe called mtcars which is constructed on the same basis as the R one.

```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"}
import pandas as pd 
mtcars = pd.read_csv("https://vincentarelbundock.github.io/Rdatasets/csv/datasets/mtcars.csv", index_col=0)
```


### Selecting rows {-}

#### Selecting row by numeric index {-}

**`R`**
```{r eval=FALSE}
mtcars[1,]  #Select row number 1
mtcars[c(1,2),] #Select row numbers 1 and 2
```

This returns a `'data.frame':	1 obs. of  11 variables` or a `'data.frame':	2 obs. of  11 variables`

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
mtcars.iloc[0,:]  #Select row number 1 (zero based indexing)
mtcars.iloc[[0,1],:]  #Select rows number 1 and 2.
```

This returns a `pandas.core.series.Series` if you select a single row, or a `pandas.core.frame.DataFrame` if you select multiple rows.

#### Selecting row by named index {-}

**`R`**
```{r eval=FALSE}
mtcars['Mazda RX4', ]
mtcars[c('Mazda RX4','Mazda RX4 Wag'), ]
```

This returns a `'data.frame':	1 obs. of  11 variables` or a `'data.frame':	2 obs. of  11 variables`

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
mtcars.loc['Mazda RX4',:]
mtcars.loc[['Mazda RX4', 'Mazda RX4 Wag'],:]
```

This returns a `pandas.core.series.Series` if you select a single row, or a `pandas.core.frame.DataFrame` if you select multiple rows.

### Selecting columns {-}

#### Selecting column by numeric index {-}

**`R`**
```{r eval=FALSE}
mtcars[,1 ]
mtcars[,c(1,2)]
```

If we select a single column, this return an atomic vector.
If we select multiple columns, it returns a `'data.frame':	32 obs. of  2 variables:`

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
mtcars.iloc[:,0]
```

If we select a single column, this returns a `pandas.core.series.Series`.
If we select multiple columns, it returns a `pandas.core.frame.DataFrame`.

### Selecting individual 'cells' (items) {-}

#### Selecting item by numeric index {-}

**`R`**
```{r eval=FALSE}
mtcars[1,1 ]
```

This returns a numeric vector with a single item

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
mtcars.iloc[0,0]
```

This returns a float

#### Selecting item by named index {-}

**`R`**
```{r eval=FALSE}
mtcars['Mazda RX4','mpg']
```

This returns a numeric atomic vector with a single item.

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
mtcars.loc['Mazda RX4','mpg]
```

This returns a float.

### Selecting items with a logical vector {-}

It is possible to use a logical vector (a series of ['truthy' and 'falsey'](https://gist.github.com/jfarmer/2647362) values) to filter a dataframe.

For instance, if your rows are `1,2,3` a logical vector `true, false, true` will select rows 1 and 3.

**`R`**
```{r eval=FALSE}
# The following df has 3 rows and 3 columns
df <- read.csv("https://gist.githubusercontent.com/RobinL/dae60170438c5e0adcd0c68fbaa6abdf/raw/62747b776c63c9bd367bcdc11fe7d6a3459eaa12/tiny.csv", row.names = 1)
logical_vector <- c(TRUE,FALSE, TRUE)
df[logical_vector,]
df[,logical_vector]

logical_vector <- c(TRUE, FALSE)
df[logical_vector,] #This works due to R's vector recyclig 
df[,logical_vector]
````

For more on vector recycling in R, see [here](http://stackoverflow.com/questions/9335099/implementation-of-standard-recycling-rules).

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
df = pd.read_csv("https://gist.githubusercontent.com/RobinL/dae60170438c5e0adcd0c68fbaa6abdf/raw/62747b776c63c9bd367bcdc11fe7d6a3459eaa12/tiny.csv", index_col=0)
logical_vector = [True, False, True]
df.loc[logical_vector,:]
df.loc[:,logical_vector]

logical_vector = [True, False]
df.loc[logical_vector,:] #Different behaviour to R - no recycling
df.loc[:,logical_vector]
```


#### Obtaining the index itself {-}

**`R`**
```{r eval=FALSE}
names(mtcars)
rownames(mtcars)
```
These are both atomic vectors of strings.

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
mtcars.index
mtcars.columns
```
These are both of type `pandas.indexes.base.Index`

###  Edge cases {-}

Things can get ambiguous in two situations:

1. You have integer row or column names which aren't just ascending integers starting at 0 (Python) or 1 (R).  This creates an ambiguity between indexing my name or by number
2. You have an index with duplicate values.


#### Integer row/columns names which aren't just 1,2,3...,n {-}

In R, row names are always strings, which resolves this ambiguity  
Data for these examples are [here](https://gist.github.com/RobinL/7378abcb98d70f642a4047f4e01fd428)

**`R`**
```{r eval=FALSE}
df <- read.csv("https://gist.githubusercontent.com/RobinL/7378abcb98d70f642a4047f4e01fd428/raw/5e26ef7581804308cd244a499737ead03eba66e8/f.csv", row.names = 1)
df[c(1,2,3),] # Selects rows in positions 1,2,3 
df[c(1,2,3,7),] # Selects rows in positions 1,2,3 and a blank row
df[c("2","4","6"),] #  #Selects rows named "2","4", and "6"" corresponding to position 1, 2 and 3
df[c("2","4","5","6"),] # Row "5" does not exist, so it creates a blank row in the results
```

**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
df = pd.read_csv("https://gist.githubusercontent.com/RobinL/7378abcb98d70f642a4047f4e01fd428/raw/5e26ef7581804308cd244a499737ead03eba66e8/f.csv", index_col=0)
df.iloc[[0,1,2],:] #Selects rows in position 1, 2 and 3
df.iloc[[0,1,2,6],:] #IndexError: positional indexers are out-of-bounds

df.loc[[2,4,6],:] #Selects rows named 2,4, and 6 corresponding to position 0, 1 and 2
df.loc[[2,4,6,99],:] #Selects rows named 2,4, and 6 corresponding to position 0, 1 and 2, plus creates a row of NA with index 99
```

You can also use the pandas function `ix` for this, but I advise against it.  See advice on the use of `ix` [here](http://stackoverflow.com/questions/31593201/pandas-iloc-vs-ix-vs-loc-explanation/31593712#31593712).

#### Duplicated row or column names {-}

Consider [this](https://gist.github.com/RobinL/05eea0870961a80814da5ae22f25f407) table.

**`R`**
```{r eval=FALSE}
df <- read.csv("https://gist.githubusercontent.com/RobinL/05eea0870961a80814da5ae22f25f407/raw/bd6f6cd2cf8edd22763e2afa460bf9a7c9bdd79b/tiny_dup.csv", row.names=1)
#Error in read.table(file = file, header = header, sep = sep, quote = quote,  : 
#  duplicate 'row.names' are not allowed
```


**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
df = pd.read_csv("https://gist.githubusercontent.com/RobinL/05eea0870961a80814da5ae22f25f407/raw/bd6f6cd2cf8edd22763e2afa460bf9a7c9bdd79b/tiny_dup.csv", index_col=0, mangle_dupe_cols=False)

df.loc[2, "b"] # Returns 2 rows and 2 columns
```

### Writing to a dataframe {-}

#### Writing to individual 'cells {-}

In what follows I use name based indexing, but the following would also work with numeric (positional) indexing.

**`R`**
```{r eval=FALSE}
#Edit existing value
mtcars["Mazda RX4", "mpg"] = 21.1

#Create new value
mtcars["Future car", "mpg"] = 999 #A new row is created.  All values are set to NA except for 'mpg' column

mtcars["Future car", "mpg"] = "Not known"  #Note that this has the effect of coercing all the other values in the 'mpg' column into a string vector.

```


**`Python`**
```{python eval=FALSE, engine.path="/Users/robinlinacre/anaconda/bin/python.app"} 
#Edit existing value
mtcars.loc["Mazda RX4", "mpg"] = 21.1

#Create new value
mtcars.loc["Future Car", "mpg"] = 999 #Note that this has the effect of coercing all the other values in the 'mpg' column into a string vector.

mtcars.loc["Future Car", "mpg"] = "Not known" #This does not affect the other values in the 'mpg' column


```

In the above I use ∞. This is the unicode string symbol for infinity.

Note that in reality, R and Python both have a numeric concept of infinity:

- R has [inf](https://stat.ethz.ch/R-manual/R-devel/library/base/html/is.finite.html).
- Python has [np.inf](https://docs.scipy.org/doc/numpy/reference/generated/numpy.isinf.html)

### Subsetting {-}


##Setup {-}

Here are some notes on running the code in this .Rmd.

To run the code, you'll want to remove `eval=FALSE` from the code blocks.

In what follows we are going to use the `mtcars` dataset.  If you want to run this code in Python, you'll need access to the dataset.  You can get this by doing:  `pip install ggplot` and then in python `from ggplot import mtcars`

Note that in order for the Python code to execute, you need to make sure that `Sys.which('python')` returns the path for Anaconda rather than Mac OS default python.  See [here](http://rmarkdown.rstudio.com/authoring_knitr_engines.html), [here](http://stackoverflow.com/questions/27097169/why-r-is-calling-system-python-instead-of-default-python-in-user-path) and [here](https://support.rstudio.com/hc/en-us/community/posts/207644217-Rstudio-server-doesn-t-have-login-user-s-environment-variables-).  The solution I found was 

````
cd ~
echo .Renviron >> PATH=/Users/robinlinacre/anaconda/bin
````


